Current Affairs: Utilizing Oxford Nanopore Sequencing Data to Detect Non-Canonical DNA Structures

### 2024-01-22 jan30meeting_pres1/may30ccgmeeting.tex
update to jan30 pres

%%%FOR VIDEOS
\usepackage{media9}
\usepackage{multimedia}
\usepackage{graphicx}
% For the Flowchart
%\usepackage{tikz}
%\usetikzlibrary{shapes.geometric, arrows}
%\tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=blue!30]
%\tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=orange!30]
%\tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=green!30]
%\tikzstyle{arrow} = [thick,->,>=stealth]

@@

% Title page details: 
\title[BEAP Dec 2022]{Double Trouble: Understanding Sex Differences in Synthetic Lethal interactions in Human Cancers \\~\\ Project Updates} 
\author{Alexander Turco}
\date{July 18, 2023}
\logo{\includegraphics[height=0.5cm, width=3cm]{logo.png}}

@@

	% For my Background slides, I will talk about important things such as WHat LCRs are, their evolution, stuff like that
	% SLIDE 3 - WHAT ARE LCRs%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

	%Also called synthetic sick because it can lead to reduced cell viability as opposed to death
	\begin{frame}{Recall: The Combined Inactivation of Two Genes Can Lead to Synthetic Lethal Interactions}
		\label{link1}

@@

			\item Inactivation: Preventing or disabling normal function of a gene (e.g mutation)
			\end{itemize}
			\centering
			\large \textbf{Synthetic = Combining}
			\footnotetext[2]{\tiny\cite{lee2018harnessing}}
	\end{frame}

@@

	%Slide 6
	% Poly(ADP-ribose) polymerases (PARPs) repair DNA SSBs through the BER pathway. PARP inhibitors, such as olaparib, prevent repair by trapping the inactivated PARP onto the SSB, resulting in the generation of DNA DSBs during the replication process. In tumors with a homologous recombination deficiency (HRD), such as a BRCA1/2 mutation, the low-fidelity repair mechanism of NHEJ leads to increasing genetic instability and ultimately death of the tumor cell.
	\begin{frame}{Recall: Synthetic Lethal Interactions are Harnessed for Precision Oncology}
		\begin{itemize} %(Olaparib, Niraparib, Rucaparib, Talazoparib)
			\item \small Four FDA approved anti-cancer drugs  are Poly [ADP-ribose] polymerase 1/2 (PARP1/2) inhibitors that work via a synthetic lethal mechanism

			\centering
			\includegraphics[width=8.5cm, height=6cm]{olaparib2.png}

			\footnotetext[4]{\tiny Figure From  \cite{o2015targeting}}
		\end{itemize}

@@

	%EXPLAIN THE PROCESS OF ISLE HERE
	\begin{frame}{Recall: Building Pan-Cancer Synthetic Lethality Networks}
		\centering
		\includegraphics[width=13cm, height=7cm]{isle.png}
		%\movie[width=10cm,height=7cm,showcontrols=true]{\includegraphics[width=10cm, height=7cm]{pompe_thumbnail.png}}{pompe_trimmed.mp4}
		\footnotetext[2]{\tiny\cite{lee2018harnessing}}
	\end{frame}

	\begin{frame}{Recall: Sex Differences Add an Additional Layer of Complexity}
		Human sex differences are mainly caused by;
		\begin{enumerate}
			\item Gonadal hormone secretions
			\item Genes located on the sex chromosomes (X and Y) \newline
		\end{enumerate}
		This leads to differences in the frequency of certain cancer types and the efficacy of treatments in males and females
	\end{frame}

	\section{Approach}
	\begin{frame}{The Objective}
		\begin{center}
			%\includegraphics[width=11cm, height=6cm]{types_of_courses_taken.jpg}
			Can we build sex-specific synthetic lethality networks for various cancer types? \newline

@@

	%%%%% APPRACH SECTION BEGINS HERE, 2 SLIDES

	\begin{frame}{Overall Project Workflow}
		\includegraphics[width=12cm, height=5.2cm]{img2.png}
	\end{frame}

	\section{Data Collection}
	\begin{frame}{Selection of TCGA Cancer Types}
			\begin{columns}
				\begin{column}{0.45\textwidth}
					\includegraphics[width=5cm, height=8cm]{img3.png}
				\end{column}
				\begin{column}{0.5\textwidth}
					\begin{itemize}
						\item There is a lack of healthy tissue RNA-seq samples in the TCGA database (12 cancers with more than 10 pairs in M and F)
						\item Raw count pan-cancer matrices created with 12 remaining TCGA cancer types (BLCA, COAD, ESCA, HNSC, KICH, KIRC, KIRP, LIHC, LUAD, LUSC, STAD, THCA)
					\end{itemize}
				\end{column}
			\end{columns}			
	\end{frame}

	\begin{frame}{Creating and Pre-Filtering Pan-Cancer Expression Matrices}
		\begin{itemize}
			\item For each gene, if expression is $>$ 90th quantile of overall expression in AT LEAST 1 sample, we keep it, otherwise we filter it out
			\item We reduce the amount of genes with extremely low expression, thereby reducing noise and improving sensitivity to detect differentially expressed genes (genes with weak expression are more susceptible to technical noise arising from library size, library composition, etc)
		\end{itemize}
	\end{frame}

	%Both samples are sequenced at the same depth - sample A has 5 genes expressed at same level, Sample B has the same genes as sample A plus two highly expressed genes that take up 50% of the reads
	\section{Data Processing}
	\begin{frame}{Normalization for RNA-seq Data Analysis}
		\centering Required to identify genes that are differentially expressed due to some biological phenomena and not due to technical variation

		\includegraphics[width=9cm,height=7cm]{img4.png}
	\end{frame}

	\begin{frame}{Normalizing Pan-Cancer Expression Matrices}
		\includegraphics[width=12cm, height=8cm]{all_cancers-F_before_and_after_normalization_v.png}
	\end{frame}

	\section{Data Processing}
	\begin{frame}{Normal and Tumor Tissue Differ ACROSS 12 Cancer Types}
		\includegraphics[width=12cm, height=6cm]{all_cancerspca_analysis_condition_0.9.png}

		Across all 12 cancer types, the observed variation between healthy and tumor tissue samples is unlikely to have occurred by chance alone (NPManova p-val = 0.0001)
	\end{frame}

	\begin{frame}{Normal and Tumor Tissue Differ WITHIN each Cancer}
		\includegraphics[width=12cm, height=6cm]{all_cancerspca_analysis_condition_cancertype_0.9.png}

		Within each cancer type, the observed variation between healthy and tumor tissue samples is unlikely to have occurred by chance alone (Acceptions: Males: ESCA (LOW SAMPLE SIZE), Females: BLCA, ESCA (LOW SAMPLE SIZE), STAD (LOW SAMPLE SIZE))
	\end{frame}

	\section{Differential Expression}
	%Our data do not fit the Poisson distribution
	%(i.e. Poisson will underestimate variability leading to an increase in false positive DE genes).
	\begin{frame}{Why does DESeq2 Use the Negative Binomial Distribution}
		\begin{itemize}
			\item Reads are count based hence they cannot be normally distributed
			\item Variance tends to be greater than the mean, especially for genes with large mean expression values
			\item  We need to account for this increase in variance using the Negative Binomial model 
		\end{itemize}
	\centering \includegraphics[width=5.5cm, height=5.5cm]{all_cancersmean_vs_variance_male.png}
	\end{frame}

	\begin{frame}{DESeq2 Accounts for Increased Variance}
		\begin{enumerate}
			\item Estimate gene-wise dispersions: captures biological variability in gene expression across samples
			\item Fit negative binomial GLM to count data: This model incorporates the estimated gene-wise dispersions as a parameter to account for variability in the data
			\item Use fitted GLM to identify differentially expressed genes \newline
		\end{enumerate}
		The use of linear models allows for more complex designs \newline \newline
		$design = ~cancertype + condition$ \newline \newline
		This tells DESeq2 to test the effect of condition while controlling for the effect of cancer type
	\end{frame}

	\begin{frame}{Finding Differentially Expressed Genes (DESeq2)}
		\includegraphics[width=12cm, height=8cm]{all_cancersvolcano_analysis_0.9.png
		}
	\end{frame}

	\begin{frame}{Differences in Number of DE Genes Males vs Females}
		\centering \includegraphics[width=9cm, height=5cm]{all_cancersvenndiagram_malede_vs_femalede_0.9.png}
		\begin{itemize}
			\item \textcolor{blue}{MALES: 358 up-regulated in tumor, 280 Down-regulated in tumor}
			\item \textcolor{yellow}{FEMALES: 171 up-regulated in tumor, 70 down-regulated in tumor}
		\end{itemize}
	\end{frame}

	\begin{frame}{DE Genes Found in Males and Females are Enriched in Cancer-Related Processes}
		\includegraphics[width=11cm, height=7cm]{all_cancersgsea_inbothdegenes_male_female_bp0.9.png}
	\end{frame}

	\begin{frame}{Enriched Biological Processes Differ in Males and Females}
		\includegraphics[width=12cm, height=8cm]{all_cancersgsea_exclusivedegenes_male_female_bp0.9.png}
	\end{frame}

	\section{Synthetic Lethality}
	\begin{frame}{Next Steps: Using DE Genes to Find Sex-Specific Synthetic Lethal Pairs}
		\begin{enumerate}
			\item Find all potential gene pair combinations using down regulated genes specific to males and specific to females \newline

			\textcolor{blue}{${280 \choose 2 } = 39060$ Downregulated Male Pairs \newline}
			\textcolor{yellow}{${70 \choose 2 } = 2415$ Downregulated Female Pairs}

			\item Check whether the co-inactivation of a candidate SL pair is underrepresented in tumor samples
		\end{enumerate}
	\end{frame}

	\section{Supplementary Slides}
	\begin{frame}{Supplementary 1 - Histogram Showing Normalization}
		\includegraphics[width=12cm, height=7cm]{all_cancers-F_before_and_after_normalization_hist.png}
	\end{frame}

	\begin{frame}{Supplementary 2 - Gene Expression Across Tissues}
		\centering \includegraphics[width=9cm, height=8cm]{all_cancerspairedboxplot_cancertype_0.9_lcal1male.png}
	\end{frame}

	\begin{frame}{Supplementary 2 - Gene Expression Across Tissues}
		\centering \includegraphics[width=9cm, height=8cm]{all_cancerspairedboxplot_cancertype_0.9_lcal1female.png}
	\end{frame}

>>>>

%%%FOR VIDEOS
\usepackage{media9}
\usepackage{multimedia}
% For the Flowchart
%\usepackage{tikz}
%\usetikzlibrary{shapes.geometric, arrows}
%\tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=blue!30]
%\tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=orange!30]
%\tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=green!30]
%\tikzstyle{arrow} = [thick,->,>=stealth]

@@

% Title page details: 
\title[BEAP Dec 2022]{Current Affairs: Utilizing Oxford Nanopore Sequencing Data to Detect Non-Canonical DNA Structures} 
\author{Alexander Turco}
\date{\today}
\logo{\includegraphics[height=0.5cm, width=3cm]{logo.png}}

@@

	% For my Background slides, I will talk about important things such as WHat LCRs are, their evolution, stuff like that
	% SLIDE 3 - WHAT ARE LCRs%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	% The graphic for this slide is an image of some output from segA with the parameters in the brackets
	\section{Background}
	\begin{frame}{A Brief History Lesson}
		\begin{itemize}
			\item (1922) American geneticist Calvin Bridges was crossing fruit flies (\textit{Drosophila melanogaster}) and observed certain genes were lethal only in combination \newline
			\item (1946) Years later, a colleague of Bridges observed the same phenomena in a different species of fruit fly, \textit{Drosophila pseudoobscura}, and coined the term `synthetic lethality' \newline
		\end{itemize}

		\centering
		\huge \textbf{Synthetic = Combining}

		\footnotetext[1]{\tiny\cite{nijman2011synthetic}}

	%Also called synthetic sick because it can lead to reduced cell viability as opposed to death
	\begin{frame}{The Combined Inactivation of Two Genes Can Lead to Synthetic Lethal Interactions}
		\label{link1}

@@

			\item Inactivation: Preventing or disabling normal function of a gene (e.g mutation)
			\footnotetext[2]{\tiny\cite{lee2018harnessing}}
		\end{itemize}

@@

	%SLIDE 5 - LCR's PRESENT IN UNIQUE WAYS
	\begin{frame}{Researchers Harness Synthetic Lethal Interactions in a Variety of Biological Applications}

		\begin{block}{\textit{To gain insight into gene function}}
			 \begin{itemize}
			 \item In \textit{Saccharomyces cerevisiae} (baker's yeast), the mapping of synthetic lethal interactions has reached a genome-wide scale and these networks of interaction are a rich source for the functional annotation of genes
			 %The elucidation of similar comprehensive synthetic lethal interaction networks in human cells would be very attractive, particularly as a large number of genes in the human genome have still not been defined a function
			 \end{itemize}
		\end{block} \pause

		\begin{block}{\textit{To uncover new opportunities for precision oncology}}
			\begin{itemize}
				\item An emerging strategy is to identify potential synthetic lethal interactions between tumor suppressor genes and other genes that can be simultaneously disrupted leading to cancer cell death (e.g. Olaparib was the first drug to work via a synthetic lethal mechanism)
				%Rather than attempting to restore function of mutated tumor suppressor genes (TSGs) which proves extremely difficult to do,
				%Talk about BRCA1 and PARP inhibitors for olaparib lol
			\end{itemize}
		\end{block}

	\footnotetext[3]{\tiny\cite{nijman2011synthetic,shen2018synthetic}}

@@

	%Slide 6
	% Poly(ADP-ribose) polymerases (PARPs) repair DNA SSBs through the BER pathway. PARP inhibitors, such as olaparib, prevent repair by trapping the inactivated PARP onto the SSB, resulting in the generation of DNA DSBs during the replication process. In tumors with a homologous recombination deficiency (HRD), such as a BRCA1/2 mutation, the low-fidelity repair mechanism of NHEJ leads to increasing genetic instability and ultimately death of the tumor cell.
	\begin{frame}{Four FDA Approved Anti-Cancer Drugs Work via a Synthetic Lethal Mechanism}
		\begin{itemize}
			\item Olaparib, Niraparib, Rucaparib, Talazoparib are Poly [ADP-ribose] polymerase 1/2 (PARP1/2) inhibitors
		\end{itemize}
		\centering
		\includegraphics[width=8.5cm, height=6cm]{olaparib2.png}

		\footnotetext[4]{\tiny Figure From  \cite{o2015targeting}}
	\end{frame}

	%SLIDE 6 
	\begin{frame}{Synthetic Lethality can also Help Explain Factors Related to the Tissue-Specificity of Cancer}

	\textbf{\color{blue}Tissue Specificity:} Cancers which manifest in different human tissues have different molecular, phenotypic, and epidemiological characteristics. Some aspects of this phenomena are; \newline

	\begin{enumerate}
		\item Variation in lifetime cancer risk
		\item Variation in cancer onset age
		\item Variation in the genes driving the cancer across tissue types \newline
	\end{enumerate}
		The literature has only reported the \textbf{\color{blue}number of lifetime stem cell divisions} and the \textbf{\color{blue}abnormal levels of CpG island DNA methylation} to be factors able to explain the variance in lifetime cancer risk across tissues (aside from well known global and cancer-specific factors)
	\end{frame}

	\begin{frame}{Quantifying cSL Load in Normal Human Tissue}
			\textbf{\color{blue}cSL Load:} Quantitatively captures the level of cancer synthetic lethal (cSL) gene pair co-inactivation based on gene expression data from normal human tissues
			\begin{center}
				\includegraphics[width=10cm, height=5.7cm]{csl.png}
			\end{center}
		\footnotetext[5]{\tiny Adapted From  \cite{cheng2021synthetic}}
	\end{frame}

	%%%%% APPRACH SECTION BEGINS HERE, 2 SLIDES
	\section{Approaches}

	\begin{frame}{Tissue cSL Load in Normal Tissues is Negatively Correlated with Lifetime Cancer Risk}
		\textbf{\color{blue}Tissue cSL Load:} Median cSL load value across all samples of each tissue type

		\centering
		\includegraphics[width=8cm, height=6cm]{tcl1.png}
		\footnotetext[5]{\tiny\cite{cheng2021synthetic}}
	\end{frame}

		\begin{frame}{Synthetic Lethality Increases Predictive Power of Models Associated with Tissue Specificity}
		\begin{columns}
			\begin{column}{0.5\textwidth}
					\includegraphics[width=7cm, height=3.5cm]{tcl2.png}
					\centering
					Stem Cell Divisions
			\end{column}
			\begin{column}{0.5\textwidth}
				\includegraphics[width=6cm, height=3cm]{tcl3.png}
				\centering
				\vspace{-0.5cm}Methylation
			\end{column}
		\end{columns}	
		\centering	
	\footnotetext[5]{\tiny\cite{cheng2021synthetic}}
	\end{frame}

	\begin{frame}{Where did the Data come From?}
		A recently published reference set of genome-wide cSLs common to many cancer types was utilized
		\begin{center}
				\includegraphics[width=6cm,height=6cm]{corenet.png}
		\end{center}
	\footnotetext[2]{\tiny\cite{lee2018harnessing}}
	\end{frame}

	%EXPLAIN THE PROCESS OF ISLE HERE
	\begin{frame}{Building Pan-Cancer Synthetic Lethality Networks}
		\centering
		\includegraphics[width=13cm, height=7cm]{isle.png}
		%\movie[width=10cm,height=7cm,showcontrols=true]{\includegraphics[width=10cm, height=7cm]{pompe_thumbnail.png}}{pompe_trimmed.mp4}
		\footnotetext[2]{\tiny\cite{lee2018harnessing}}
	\end{frame}

	%%% EXPLAIN NSSE SURVEY A BIT
	%%% WHEN YOU TALK MENTION THAT THEY DID THIS SURVEY AFTER PLAYING THE GAME IN CLASS
	\begin{frame}{There is Variability in Synthetic Lethal Interactions Depending on Cancer Type}

		\textbf{\color{blue}Pan-Cancer Analysis:} Analysis of multiple types of cancer simultaneously, aims to identify commonalities shared across different types

		\textbf{\color{blue}Cancer-Specific Analysis:} Analysis of individual cancer types to better understand the variation within a specific cancer type \vspace{1cm}

		\begin{itemize}
			\item Pan-cancer analyses may overlook cancer-specific differences
			\item Availability of datasets for all cancer types varies, this may limit statistical power of pan-cancer analyses
		\end{itemize}

	%%%%RESULTS BEGIN HERE
	\section{Results}

	\begin{frame}{Sex Differences Add an Additional Layer of Complexity}
		Human sex differences are mainly caused by;
		\begin{enumerate}
			\item Gonadal hormone secretions
			\item Genes located on the sex chromosomes (X and Y) \newline
		\end{enumerate}
	This leads to differences in the frequency of certain cancer types and the efficacy of treatments in males and females
	\end{frame}

	\begin{frame}{Sex Differences Add an Additional Layer of Complexity}
		\centering
		\includegraphics[width=6cm,height=7cm]{sex.png}
		\footnotetext[6]{\tiny\cite{oliva2020impact}}
	\end{frame}

	\begin{frame}{The Objective}
		\begin{center}
		%\includegraphics[width=11cm, height=6cm]{types_of_courses_taken.jpg}
		Can we build sex-specific synthetic lethality networks for various cancer types? \newline

@@
